---
title: "Session 1: Basics in R (review)"
author: "Stephany Orjuela"
date: "November 13, 2018"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General remarks

Before you start working on a project, you should think about and decide on a directory structure. Also, if you have all your code saved in scripts, you can always rerun every step of the analysis.

Some general recommendations on working with R:

  * Refresh your R session every time you start working on something new (In RStudio, Session .. Restart R).
  * Organize your scripts into reasonable subunits (e.g., for different purposes/topics, create separate scripts)
  * Try and use a consistent style in the way that you name variables, files and functions
  * Use scripts to save code. Never save the workspace!
  * Save only the R objects that will be needed later with `saveRDS()`; these can be loaded in a later script with `readRDS()`

We will use Rmarkdown to convert an Rscript to a report that includes text, code and figures all together into an HTML report (PDF can also be used). You can find some Rmarkdown basics [here](https://rmarkdown.rstudio.com/authoring_basics.html). See the following articles for a more detailed explanation of the [code chunk options](https://rmarkdown.rstudio.com/authoring_rcodechunks.html) and [pandoc markdown](https://rmarkdown.rstudio.com/authoring_pandoc_markdown.html)


# Obtaining the data

Our data will not always be on our computer. We must be able to obtain the data from wherever it is.  We will download the data for this exercise from a github repository. To learn more about github click [here](https://guides.github.com/activities/hello-world/)

### For mac and unix users:

From your terminal, create a working directory for today's session and move to it

```{bash eval=FALSE}
mkdir BIO255
cd BIO255
```

Download the data to your computer with the following command
```{bash eval=FALSE}
github clone https://github.com/sorjuela/BIO255-Cancer-Epigenetics.git
```

### For windows users:

* Go to `https://github.com/sorjuela/BIO255-Cancer-Epigenetics` 
* Click on the `Clone or download` button, and select `Download ZIP`
* unzip the folder 

# First steps in R (quick and dirty reminder)

### Data types and assignment

```{r}
a1 <- 2
a2 <- "abc"
a3 <- TRUE
```

### Simple computation
```{r}
3+2
3*2
3^2
```

### Containers (1d)  
```{r}
## Vector: Same data types
v <- c(1,2,3)

v # !!! automatic type conversion !!!

## generate vector
v <- c(21,22,23,24)
v <- 21:24 ## less typing
v <- seq(from=21,to=24,by=1)

## select by position
v[2]
v[-2]
v[1:3]
v[c(1,4)]

## select by value
v == 22
v[v==22]
v<=22
v[v<=22]
v %in% c(21,24)
v[v %in% c(21,24)]

## list: different data types
l <- list(a="a", b=1, c=FALSE)
l
l$a
l$b
```

### Containers (2d)
```{r}
## Matrix
1:9
m <- matrix(1:9, ncol = 3, nrow = 3)
m ## wrapped around col by col
m[1,2]
m[1:2, 1:2]
m[1,]
m[,1]

## data frame
head(mtcars)
dim(mtcars)
nrow(mtcars)
ncol(mtcars)
str(mtcars)
?mtcars
mtcars$hp

dat <- mtcars
dat$cyl <- as.factor(dat$cyl)
dat$am <- factor(dat$am, levels = c(0,1), labels = c("automatic", "manual"))
summary(dat$cyl)
dat[dat$cyl == "4",]
```


### Some useful built-in functions
```{r}
cat("abc")
paste("abc", "def")
x <- "def"
paste("abc", x)
```

### Plotting
```{r}
# simple plot
x <- rnorm(100)
hist(x)
hist(x, main = "Mein Titel")
plot(mtcars$hp, mtcars$qsec)

## ggplot2
head(dat) ## mtcars

#To install:
#install.packages("ggplot2")

library(ggplot2)
ggplot(data = dat, aes(x=hp, y=qsec)) + geom_point()
ggplot(data = dat, aes(x=hp, y=qsec)) + geom_point() + aes(color = am)

## color by group
myVar <- "am"
ggplot(data = dat, aes(x=hp, y=qsec)) + geom_point() + aes(color = myVar)
ggplot(data = dat, aes(x=hp, y=qsec)) + geom_point() + aes_string(color = myVar)

## add labels
ggplot(data = dat, aes(x=hp, y=qsec)) + geom_point() + aes_string(color = myVar) + labs(x="Power", y="Time",title="bla")
```

